Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model

### 2014-04-11 paper_enahpc14/Introduction.tex
WS: significant pruning to make way for results from Manuel and Sandra

Comprehensive  scientific assessments  of climate-warming  trends over
the  past 50  years have  shown  that the  state and  dynamics of  the
Earth's climate system have  undergone unprecedented major changes, as
evidenced  by  increases  in   global  average  atmosphere  and  ocean
temperatures,  widespread melting  of  snow and  ice,  and rising  sea
levels.   The last report  of the  Intergovernmental Panel  on Climate
Change, released  in September 2013  \citep{IPCC-2013}, concluded that
most of the observed temperature  increases since the mid 20th Century
have  been caused  by  increasing concentrations  of greenhouse  gases
resulting from  anthropogenic activities  such as fossil  fuel burning
and  deforestation.

The Earth's  surface temperature is determined by  the balance between
incoming  solar radiation and  outgoing infrared  radiation. Radiative
forcing describes any imbalance  in Earth's radiation budget caused by
human  or natural interventions  in the  climate system.   It directly
measures the capacity of a gas  or other forcing agents to affect that
energy balance, thereby contributing to climate change.  The intensity
of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
emission of heat by various  materials, all differ in their warming or
cooling contributions to climate forcing.

The radiative forcing  potential of a greenhouse gas  is determined by
its  atmospheric concentration, warming  capacity, residence  time and
spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
dioxide, methane and nitrous oxide, which linger chemically stable and
homogeneously well-mixed over time scales  of a decade to thousands of
years  in the  atmosphere,  are major  human-induced climate  drivers;
whereas  short-lived gases such  as water,  suphur dioxide  and carbon
monoxide,  which are  abundant heat-trapping  gases but  physically or
chemically  reactive  to changes  in  temperature,  cause positive  or
negative  feedbacks effects which  may vary  with forcing  history and
climate  state. On a  per molecule  basis, other  gases can  have more
potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
fortunately far less abundant in the atmosphere.

One  factor   overwhelmingly  affects  the   uncertainty  in  defining
anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
the  atmosphere.  Although  they  are not  considered a  heat-trapping
greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
significantly affected the atmospheric  radiative fluxes over the last
150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
concentrations can impact the climate system by reflecting (e.g.  pure
sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
radiation  and  thereby exert  a  cooling  or  warming effect  on  the
Earth-atmosphere   system,  causing   a  so-called   direct  radiative
forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
distribution  and chemical  composition, they  can also  act  as cloud
condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
microphysical   processes  and   optical   properties  (cloud   albedo
effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
formation.      This     results     in     changes     in     droplet
concentrations                \citep{Albrecht-1989}                and
precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
aerosol  particles  represents  a  major  removal  mechanism  for  air
pollutants from  the atmosphere, however the  effects of anthropogenic
aerosols on  clouds and  the hydrological cycle  as well as  the cloud
lifetime    effect    are    especially    hard    to    assess    and
quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
uncertainties in climate modeling and in climate change prediction due
to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
Rosenfeld-2013, Lee-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  uncertainties  of aerosol  effects  in  climate, the  research
community is making a  concerted international effort to represent the
underlying chemical  processes through models.  These  models, such as
ART (Aerosols  and Reactive Trace  gases) extension, developed  at the
Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
reduce   the  climate  uncertainty,   particularly  on   the  regional
scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
be coupled with climate models,  such as the regional weather forecast
model COSMO (Consortium  for Small-scale Modelling), jointly developed
by  a consortium  of  European weather  centers  including the  German
weather service  DWD and MeteoSwiss,  and used in the  climate version
(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
model provides  a detailed description of air  pollution chemistry and
aerosol processes,  and is  mainly designed to  study air  quality and
aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
scales.  It is capable of  simulating aerosol distributions as well as
their interactions over Europe as well as other regional domains.

>>>>

Comprehensive  scientific assessments  of climate-warming  trends over
the  past 50  years have  shown  that the  state and  dynamics of  the
Earth's climate system have  undergone unprecedented major changes, as
evidenced  by  increases  in   global  average  atmosphere  and  ocean
temperatures,  widespread melting  of  snow and  ice,  and rising  sea
levels.  The suspected culprit for this change are human-induced or anthropogenic
factors, such as the greenhouse gas emissions from the burning of 
fossil fuels, which may be trapping more incoming solar radiation
in the earth system.

One  factor   overwhelmingly  affects  the   uncertainty  in  defining
anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
the  atmosphere.  Although  they  are not  considered a  heat-trapping
greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
significantly affected the atmospheric  radiative fluxes over the last
150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
concentrations can impact the climate system by reflecting (e.g.  pure
sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
radiation  and  thereby exert  a  cooling  or  warming effect  on  the
Earth-atmosphere   system,  causing   a  so-called   direct  radiative
forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
Liao-2005,  Bangert-2012,  Lundgren-2013}.

Hence to  improve our understanding of  aerosol-cloud interactions and
reduce  uncertainties  of aerosol  effects  in  climate, the  research
community is making a  concerted international effort to represent the
underlying chemical  processes through models.  These  models, such as
ART (Aerosol Reactive Transport), developed  at the
Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
reduce   the  climate  uncertainty,   particularly  on   the  regional
scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
be coupled with climate models,  such as the regional weather forecast
model COSMO (Consortium  for Small-scale Modelling), jointly developed
by  a consortium  of  European weather  centers  including the  German
weather service  DWD and MeteoSwiss,  and used in the  climate version
(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
model provides  a detailed description of air  pollution chemistry and
aerosol processes,  and is  mainly designed to  study air  quality and
aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
scales.  It is capable of  simulating aerosol distributions as well as
their interactions over Europe as well as other regional domains.

### 2014-04-11 paper_enahpc14/Sec2:COSMO-ART.tex
WS: significant pruning to make way for results from Manuel and Sandra

Aerosols dynamics  are simulated  with the modal  aerosol microphysics
module  MADE \citep{Ackermann-1998}, expanded  in MADEsoot  to feature
explicit  treatment of  soot aging  through condensation  of inorganic
\citep{Riemer-2003} and organic substances. MADE describes the aerosol
population  through  five   modes  representing  sub-micron  particles
consisting of sulphate, ammonium, nitrate, particulate organic matter,
water and soot  \citep{Riemer-2004} in a range of  mixing state. These
modes are coupled  with the gas phase by  condensation and nucleation,
and are  strongly influenced by  anthropogenic emissions of  gases and
particles. MADEsoot  describes the sub-micrometer  aerosol population,
composed  of sea  salt \citep{Lundgren-2013},  mineral  dust particles
\citep{Vogel-2006,  Stanelle-2010},   by  means  of   six  interacting
log-normal modes.  For each  mode, mass contributions and total number
concentration are prognostic  quantities, while the standard deviation
is fixed.

Specific  modules are included  to simulate  the dispersion  of pollen
grains    \citep{Vogel-2008}   and    other    biological   particles.
Meteorologically-influenced emissions  are also online  coupled within
the  model  system.  The  biogenic  VOC  (volatile organic  compounds)
emissions are  calculated as functions of  the land use  type based on
the Global  Land Cover 2000  dataset and the modeled  temperatures and
radiative fluxes \citep{Vogel-1995}.

The gaseous chemistry in COSMO-ART  is solved by a modified version of
the  Regional  Acid  Deposition  Model, Version  2  (RADM2)  mechanism
\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
organic aerosol formation based on a volatile basis set (VBS) approach
\citep{Athanasopoulou-2013}  and  hydroxyl  radical recycling  due  to
isoprene chemistry \citep{Geiger-2003}.  The thermodynamic equilibrium
between the  gas and particulate  phases of the inorganic  material is
achieved through the  ISORROPIA II module \citep{Fountoukis-2007}.

COSMO-ART  is  fully  online-coupled,  and  allows  for  feedbacks  of
aerosols  on  temperature,  radiation  and cloud  condensation  nuclei
(CCN).   Analytical  description  of  modules  incorporated  for  this
purpose  exists in  \citep{Vogel-2009,  Bangert-2011}.  The  radiation
scheme used  within the  model to calculate  the vertical  profiles of
shortwave and longwave radiative fluxes is GRAALS \citep{Ritter-1992}.
In order to account for  the interaction of aerosol particles with the
cloud microphysics and radiation,  COSMO-ART uses the two-moment cloud
microphysics  scheme of  Seifert and  Beheng  \citep{Seifert-2006} and
comprehensive parameterizations for cloud condensation and ice crystal
nucleation  \citep{Bangert-2011, Bangert-2012}.   The system  of stiff
ordinary differential equations described by chemical reactions in the
aqueous-phase together with the transfer reactions is solved using the
Kinetic PreProcessor (KPP, \citealp{Damian-2002}).

@@

Thus, we give here a brief  overview of the main features of the model
setup for  our benchmark's daily  runs.  Three-dimensional simulations
are performed over  Europe for April 13th 2010, which  is close to the
equinox and thus  brings benefits of having approximately  half day of
night  and half  day  of  sun exposure,  therefore  ensuring a  proper
activation of the chemistry cycle. We consider daily 24-hour forecasts
without  any  previous spin-up  simulations.   The calculation  domain
corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
$222\times   216$    points   with   a    horizontal   resolution   of
$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
layers.   Boundary  and  initial  conditions, emissions  and  external
parameters  were obtained by  the INT2LM  pre-processing interpolation
tool, which interpolate  data from a coarse-grid driving  model to the
rotated    latitude-longitude   grid    of   the    COSMO-Model,   and
INT2LM\_trcr\_ART,  which  interpolates  aerosol  or  gaseous  species
initial and boundary conditions  and consists mostly of the definition
of a large number of  tracers. In particular when using COSMO-ART, the
following input data are necessary:

>>>>

@@

Thus, we give here a brief  overview of the main features of the model
setup for  our benchmark's daily  runs.  Three-dimensional simulations
are performed over  Europe for April 13th 2010, which  is close to the
equinox and thus  brings benefits of having approximately  half day of
night  and half  day  of  sun exposure,  therefore  ensuring a  proper
activation of the chemistry cycle. We consider daily 24-hour forecasts
without  any  previous spin-up  simulations.   The calculation  domain
corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
$222\times   216$    points   with   a    horizontal   resolution   of
$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
layers.   COSMO-ART requires the following input data:
